Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-06 intro.tex
some wording changes

Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series\cite{weigend-book}, let alone an arbitrary time series. This naturally leads to an important and hard question: given a complicated real-valued time series, does their exist a forecast model for that time series that can utilize the information being transmitted from past to future (if such information exists)? A first step in answering this question is to answer: is it possible to reliably quantify where on the spectrum of complexity the time series lies? i.e, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) can it be inferred without a model if the signal is too \emph{complex} to effectively forecast? Since the information in an observation can be partitioned into two pieces --- redundancy and entropy generation\cite{crutchfield2003} --- our goal is to analyze how the amount of predictive structure present in a signal, as measured by the redundancy, relates to the performance of prediction strategies, as well as what is meant by \emph{low, medium and high} complexity in Section \ref{sec:meaComplex}.

To answer this question and for the purposes of this paper we focus on real-valued (possibly) noisy scalar time series which appear \emph{complicated}, i.e., they exhibit interesting behavior, for instance, not of low period, strictly monotonic, or constant. We make no assumptions about the generating process's properties,(e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary, etc.) and we attempt to measure empirically where on the previously mentioned spectrum of \emph{complexity} the time series exists.
We argue that an extension of \emph{permutation entropy}
\cite{bandt2002per}---a method for approximating the entropy of a real-valued-finite-length-noisy time series through ordinal analysis---is an effective way to explore that conjecture. In light of this, for the purposes of this paper we define \emph{complexity} as an approximation of Shannon \emph{entropy}, which we define and explain rigorously in Section \ref{sec:meaComplex}. In this fashion a random walk time series (which exhibits high entropy) is purely complex, where as a periodic signal (which exhibits low entorpy) has low complexity. The permutation entropy is ideal in a situation like this as it converges on the true entropy value, unlike other partitioning techniques used in the past which either require specific knowledge of the generating process or produce biased values of the entropy~\cite{bollt2001}.

>>>>

Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series\cite{weigend-book}, let alone an arbitrary time series. This naturally leads to an important and hard question: given a complicated real-valued time series, does their exist a forecast model for that time series that can utilize the information being transmitted from past to future (if such information exists)? A first step in answering this question is to answer: is it possible to reliably quantify where on the spectrum of complexity the time series lies? i.e, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) can it be inferred without a model if the signal is too \emph{complex} to effectively forecast? Since the information in an observation can be partitioned into two pieces --- redundancy and entropy generation\cite{crutchfield2003} --- our goal is to analyze how the amount of predictive structure present in a signal, as measured by the redundancy, relates to the performance of prediction strategies. The formation of this spectrum or partition as well as what is meant by \emph{low, medium and high} complexity are visited in more detail in Section \ref{sec:meaComplex}.

To answer this question and for the purposes of this paper we focus on real-valued (possibly) noisy scalar time series which appear \emph{complicated}, i.e., they exhibit interesting behavior, for instance, not of low period, strictly monotonic, or constant. We make no assumptions about the generating process's properties,(e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary, etc.) and we attempt to measure empirically where on the previously mentioned spectrum of \emph{complexity} the time series exists.
We argue that an extension of \emph{permutation entropy}
\cite{bandt2002per}---a method for approximating the entropy of a real-valued-finite-length-noisy time series through ordinal analysis---is an effective way to explore that conjecture. In light of this, for the purposes of this paper we define \emph{complexity} as a particular approximation of Kolmogorov-Sinai \emph{entropy}[[Probably need a citaiton for KS entropy]], which we define and explain rigorously in Section \ref{sec:meaComplex}. In this fashion a random walk time series (which exhibits high entropy) is purely complex, where as a periodic signal (which exhibits low entorpy) has low complexity. The permutation entropy is ideal in a situation like this as it converges on the true entropy value, unlike other partitioning techniques used in the past which either require specific knowledge of the generating process or produce biased values of the entropy~\cite{bollt2001}.

### 2014-03-06 wpe.tex
some wording changes

Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
it is common in the literature to normalize permutation entropy as follows:
$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
and ``high'' entropy is close to 1. Finally, it should be noted that the
permutation entropy has been shown to be identical to the Shannon entropy for
many large classes of systems \cite{amigo2012permutation}.

>>>>

Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
it is common in the literature to normalize permutation entropy as follows:
$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
and ``high'' entropy is close to 1. Finally, it should be noted that the
permutation entropy has been shown to be identical to the Shannon entropy for {\color{red}[[Ryan is this the case or is it the KS entropy?]]}
many large classes of systems \cite{amigo2012permutation}.
